---
title: "Improving the Transparency of Power Analyses Through Standardized Reporting"
author:
  - name: Cristian Mesquida
    orcid: 0000-0002-1542-8355
    email: c.mesquida.caldentey@tue.nl
    affiliations:
      - name: Eindhoven University of Technology
        department: Industrial Engineering and Innovation Sciences
        address: Den Dolech 2, 5600 MB, Eindhoven
        city: Eindhoven
    roles:
     - conceptualization
     - methodology
     - formal analysis
     - data curation
     - writing — original draft
  - name: Daniël Lakens
    corresponding: true
    orcid: 0000-0002-8393-5316
    email: D.Lakens@tue.nl
    affiliations:
      - name: Eindhoven University of Technology
        department: Industrial Engineering and Innovation Sciences
        address: Den Dolech 2, 5600 MB, Eindhoven
        city: Eindhoven
    roles:
     - conceptualization
     - methodology
     - formal analysis
     - data curation
     - writing — review and editing
  - name: Lisa DeBruine
    orcid: 0000-0002-7523-5539
    email: lisa.debruine@glasgow.ac.uk
    affiliations:
      - name: University of Glasgow
        department: Institute of Neuroscience and Psychology
    roles:
     - conceptualization
     - methodology
     - formal analysis
     - data curation
     - writing — review and editing
keywords: [Peer Review, Quality Control, Automation, Standardisation, Power Analysis]
author-note:
  disclosures:
    conflict-of-interest: "Authors declare no conflicts of interest"
    data-sharing: "The data and analysis scripts related to this study are publicly available on the Open Science Framework and can be found at XXX"
bibliography: [references.bib, package-citations.bib]
floatsintext: true # Place figures and tables within the text
format:
  apaquarto-html:
    embed-resources: false
    toc-depth: 3
    toc-expand: true
    css: style.css
  apaquarto-docx: default
execute:
  echo: false
abstract: ""
---

```{r}
#| label: setup
#| eval: FALSE

# install the apaquarto extension
quarto::quarto_add_extension("wjschne/apaquarto", no_prompt = TRUE)

# install the main version of Metacheck
pak::pkg_install("scienceverse/metacheck@main")
```

```{r}
#| label: libraries
#| message: FALSE
#| output: FALSE

# Libraries to be loaded
library(metacheck)
library(dplyr)
library(tidyr)
library(flextable)

# Create BIB file for R packages
knitr::write_bib(
  .packages(),
  file = "package-citations.bib"
)
```

```{r}
#| label: data

# Load RDS object containing articles from 2014 to 2026
psych_science <- readRDS("data/psychsci.rds")
```

## Introduction

An important step when designing a study is to justify the sample size that will be collected. The key aim of a sample size justification is to explain how the collected data is expected to provide valuable information given the inferential goals of the researcher [@lakens2022]. One common sample size justification when the inferential goal is to test a hypothesis is an *a priori* power analysis. An a priori power analysis makes explicit which effect size researchers are interested in detecting. This information is essential both when interpreting the informational value of a study, as when interpreting the test results. A study that has high power only for unrealistically large effect sizes is unlikely to contribute useful insights to the published literature. If a hypothesis test yields a non-significant result, the range of effect sizes that could have been detected with low error rates is made explicit by the power analysis, which provides important context to evaluate the test results. For these reasons, researchers need report all parameters used to compute the power analysis. 

Despite the important role power analyses play in hypothesis testing research, there are no reporting guidelines for power analyses. Furthermore, when a priori, sensitivity, or compromise power analyses are reported, researchers often omit important information [@thibault_2022; @thibault_errors_power; @bakker_2020]. This lack of transparency prevents editors, reviewers, and readers from being able to critically evaluate the validity of the power analysis, and to reproduce its results. Metascientific research has observed discrepancies between the statistical test specified in the *a priori* power analysis and the key statistical test performed in the study [@bakker_2020]. [CRIS< THIS NEEDS A BIT MORE EMPIRICAL WORK ON POWER ANALYSES AND THEIR PROBLEMS. INCLUDING YOUR OWN PAPER NOT REPRODUCING THEM], or between the effect size used in the power analysis and the effect of interest in a newly conducted study.

To improve the way power analyses are reported in the scientific literature, we propose a four-step solution. First, reporting guidelines need to be developed that provide clear and universal guidance on how to transparently and completely report power analyses. Second, scientist need to be educated about how to follow these reporting guidelines. Third, tools should be developed that assist researchers by automatically checking whether they are reporting power analyses in line with reporting standards. And fourth, peer reviewers and editors should be able to easily check whether researchers report power analyses correctly. In this manuscript we aim to contribute to all four steps of this solution by creating reporting guidelines for power analyses, and subsequently incorporating these reporting guidelines in a Metacheck module automatically checks whether power analyses are accurately reported. This Metacheck module can educate researchers about best practices, assist researchers in improving how they report power analyses, and allows editors and reviewers to quickly check if manuscripts follow best reporting practices.  

#### Reporting Guidelines for Power Analyses

One reason why power analyses lack key information is the absence of specific reporting guidelines that editors, reviewers, and researchers can follow. For example, the American Psychological Association's (APA) *Journal Article Reporting Standards* (JARS) provides reporting standards for several aspects of scientific articles [@appelbaum2018], such as p-values and confidence intervals, but it provides no guidance on how to report power analyses. Furthermore, JARS reporting guidelines specify *what* information to report, but rarely provide detailed information about *how* the information should be presented. For example, JARS for quantitative articles state that researchers should describe the "power analysis, or methods used to determine precision of parameter estimates" but does not specify how this information should be reported. Given the importance of power analyses for the evaluation of research findings, a reporting standard that specifies both the required information and a consistent structure for how this information should be presented could substantially improve current research practices.

A useful precedent exists in the standardized reporting of statistical tests following APA guidelines. Researchers are familiar with reporting statistical tests such as *t*-tests and *F*-tests in a consistent order that includes the test statistic, degrees of freedom, *p*-value, effect size, and it's confidence interval. For example, an *F*-test may be reported as *F*(1, 3) = 3.23, *p* = .01, $\eta_p^2$ = .12 [**Cris, add a CI**]. These standardized reporting convention allows readers to quickly identify and interpret the information they convey. At the same time, standardized reporting practices have the important benefit that they enable automated extraction of tests, which is used by tools such as Statcheck to reliably parse statistical tests and verify reported *p*-values [@nuijten_prevalence_2015].

The same approach can be applied to power analyses. The first step is to determine what information should be reported and how it should be reported. A fully reported power analysis requires an identifier (as there are often multiple power analyses in a paper), the type of power analysis (i.e., a priori, sensitivity, compromise, post-hoc), exact details about the statistical test that the power analysis is performed for, details about the effect that the power analysis aim to detect, and the Type 1 and Type 2 error rates used in the calculation. As different software packages can give slightly different results, the software used to compute power should be reported as well. Combining these properties in our proposed standard report for power analyses would leave to a sentence such as:

An a priori power analysis in G*Power for a 2wx2w F-test for an interaction effect, assuming a within-participant r = 0.5, to detect f = 0.12, based on the previously published study by Someone et al., (2025), with a desired power of 0.90 and an alpha of 0.05, returned a sample size of 186 observations.

Alternatively, and just as statistical tests are abbreviated instead of communicated in prose, one could write: 

A power analysis (a priori, G*Power, F-test, 2wx2w, interaction effect, assuming within-participant r = 0.5, f = 0.12, previous study (Someone et al., 2025), Type 1 = 0.05, Type 2 = 0.20, www.figshare.com/screenshot_of_gpower) indicated that 186 observations should be collected. 

Generically, the format for an a priori power analysis is ordered as follows: Type of power analysis, software, test family, test parameters, effect size, effect size justification, alpha level, desired power, link to reproducible file. For other types of power analyses the achieved sample size is added, and the computed variables are moved to the end of the statement. For example, a sensitivity power analysis based on a simulation in R would be reported as:

A power analysis (sensitivity, R, t-test, two-sample, N = 90 (45 per group), Type 1 = 0.05, Type 2 = 0.10, www.data_repository.org/link_to_simulation_script) indicates that effects of d = 0.69 could be detected with the desired 90% power.

[Cris, the above is better than the text below, right? More in line with APA. What is more difficult is that there are many more possible text entries - maybe we need to make an app that allows them to pick from a dropdown menu.]
> power_analysis: id = 01; type = a priori; family_test = F-test; type_test = 2w2w; test_parameters.within_subject_correlation = 0.50; type_effect = interaction; effect_size.metric = f; effect_size.magnitude = 0.12; effect_size_justification.method = previous study; effect_size_justification.reference = https://doi.org/example/251524592097094; sample_size.total = 186; alpha.level = 0.05; power = 0.90; software.name = G\*Power, software.file_location = https://EXAMPLE".


## A Structured Schema for Power Analyses

To formalize these reporting requirements and facilitate power analyses to be machine-readable, we have developed a JSON schema that defines the parameters to be reported. The schema contains the data types, permitted values, and any relationships or conditional requirements between characteristics of a reported power analysis. The JSON schema serves as a formal machine-readable specification of the reporting standard. Researchers are not expected to report power analyses in this format, but to report power analyses clearly enough such that software can parse their writing, and add it to a schema automatically (and without the use of large language models). An example of the full .JSON file is presented below, and is followed by a detailed explanation of each aspect of a power analysis that should be provided for full reporting.

Box 1: Example JSON representation of a complete power analysis object.

``` json
{
  "id": "01",
  "type": "a priori",
  "family_test": "F-test",
  "type_test": "2w*2w",
  "test_parameters": [
    "within_subject_correlation: 0.5"
  ],
  "type_effect": "interaction",
  "effect_size": {
    "metric": "f",
    "magnitude": 0.12
  },
  "effect_size_justification": {
    "method": "previous study",
    "reference": "https://doi.org/example/251524592097094"
  },
  "sample_size": {
    "total": 186
  },
  "power": 0.9,
  "alpha": {
    "level": 0.05
    },
  "software": {
    "name": "G*Power",
    "file_location":"https://EXAMPLE"
  }
}
```


<!--#  The cases included for each component are not intended to be exhaustive. They were deliberately simplified to illustrate the proposed reporting framework. In practice, different types of power analyses may require different schemas, or alternatively, a single schema with conditional fields that depend on the type of power analysis. -->

### Unique Identifier

We propose that each power analysis should be assigned a unique identifier. This identifier creates an explicit link between the research question, the statistical test used to address it, and the corresponding power analysis. Power analyses are often reported with insufficient detail to determine which hypothesis or statistical test they were intended to support [Cris, cite your own paper]. 

This issue is particularly relevant in studies that include multiple hypotheses or multiple statistical tests. Assigning a unique identifier to each power analysis establishes these relationships and determine whether each planned analysis is supported by an appropriate power analysis. Such links could facilitate automated checks of whether all primary analyses have a corresponding power analysis and whether multiple power analyses refer to the same planned analysis. The `id` field should contain a unique identifier that allows the power analysis to be linked to the corresponding hypothesis and statistical test. Whee researchers often list and number hypothesis (e.g., H1, H2, H3), it should be just as normal to list multiple power analyses, as each hypothesis corresponds to a unique test, with it's own statistical power for an effect size of interest. 

[ I would not show all these json files in the main text - 1 complete example is enough. ]
Box 3.1: JSON Schema representation of the `id` property.

``` {#id .json}
{
  "id": {
    "type": "string",
    "description": "Unique identifier used to link the power analysis to the corresponding hypothesis and statistical test."
  }
}
```

Box 3.2. Example JSON representation of the `id` property.

``` json
{
  "id": "01",
  ...
}
```

### Type of power analysis

There are four types of power analyses. First, an *a priori* power analysis calculates the sample size required to detect an effect of interest, and is conducted before data collection. Second, a sensitivity power analysis calculates the effect size that can be detected with desired statistical power given the available sample size, statistical test, and significance level. [Cris, I would just refer to it as sensitiviy, and say that a 'correct' post-hoc is also sensitivity, in essence. In the schema, we refer to this type of analysis as effect-size sensitivity analysis to distinguish it from other forms of sensitivity analyses.] Third, researchers can perform a compromise power analysis where they compute the desired Type 1 and Type 2 error rates, given the sample size, and the effect size of interest. Finally, a post hoc power analysis calculates the statistical power associated with a prespecified effect size given the sample size actually obtained, and the alpha level.

It is important to point out that the term post hoc power analysis is frequently used ambiguously. Researchers may use it either to refer to power calculated based on a hypothetical effect size of interest, which can provide information about the sensitivity of the study design given the sample size. Researchers can also use it to calculate power using the effect size observed in a dataset, which provides no additional information beyond the *p*-value [@yuan_2005; @lakens2022]. Researchers have recommended to avoid the use of 'post hoc' power, [@okeefe_brief_2007]. There is no term used for the correct use of a power analysis where the power of a test is computed for an effect size of interest. We propose to use *hypothetical power analysis* when computing the power of a test given an effect size of interest, as opposed to *observed power analysis* where power is computed for the observed effect size. 

Box 4.1: JSON Schema representation of the `type` property.

``` json
"type": {
      "type": "string",
      "description": "The type of power analysis conducted.",
      "enum": [
        "a priori", 
        "effect size sensitivity", 
        "power sensitivity",
        "post hoc",
        "comprimise"
      ]
    }
```

Box 4.2: Example JSON representation of the `type` property.

``` json
{
  ...
  "type": "a priori",
  ...
}
```

<!--# I assume it is better to avoid spaces between words. -->
<!--# Indeed - I would make all labels connected a_priori - DL. -->

### Test Family and Test Type

Researchers should provide both general information about the family the test that power is computed for belongs to (e.g., a correlation, *t*-test, or *F*-test) as more specific information about the type, design, or configuration of that test. In essence, this aspect of the power analysis contains all information that would be specified in a power analysis about the statistical test that is performed. For example, for a *t*-test, power analysis software such as G\*Power would require users to specify the type of *t*-test (one-sample, two-sample, or paired-sample). 

For *F*-tests, researchers should specify the factorial structure (for example using notation such as 2w, 2b, or 2b\*2w, where the number indicates the number of levels and the letters indicate within-subject (w) or between-subject (b) factors) and the which main effect or interaction the power analysis is for. The effect of interest should be specified because a statistical test does not always unambiguously identify the specific effect for which a power analysis was conducted. For simple designs, such as a *t*-test, the effect of interest typically corresponds directly to the comparison being tested. However, in factorial designs, researchers may be interested in different effects, including main effects, interactions, planned contrasts, or simple effects. Specifying the type of effect clarifies which hypothesis the power analysis was designed to evaluate and allows readers to determine whether the power analysis corresponds to the intended research question.

Some statistical tests require additional parameters to fully specify the power analysis. These test parameters depend on the test family and the design configuration. Test assumptions (e.g., the homogeneity of variances assumption that differentiates Student's *t*-test from Welch's *t*-test, or a Greenhouse Geisser correction for an *f*-test) are one example of test parameters. For correlations and *t*-tests the directionality of the hypothesis should be reported using the tail parameter. For within-subject *F*-tests the correlation between repeated measurements should be reported. This correlation represents the expected association between repeated observations and influences the variance estimate used in the power calculation. In complex designs, this would require specifying a correlation matrix. The more complex a test is, the more details of a test need to be specified, up to a point where it might be easier to share the code for the power analysis than to describe all details of the test verbally.

Box 5.1: JSON Schema representation of the `family_test` property.

``` json
"family_test": {
      "type": "string",
      "description": "Broad class of statistical test used in the power analysis.",
      "enum": [
        "correlation", 
        "t-test", 
        "F-test"
        ]
    }
```

Box 5.2: Example JSON representation of the `family_test` property.

``` json
{
  ...
  "family_test": "F-test",
  ...
}
```

Box 6.1: JSON Schema representation of the `type_test` property.

``` json
"type_test": {
      "type": "string",
      "description": "Specific test configuration. For r, report 'within-subject' or 'between-subject'. For t-tests, report 'one-group', 'within-subject', or 'between-subject'. For F-tests, use factorial notation such as 2w, 2b, or 2b*2, where the number indicates the number of levels and the letters indicate between-(b) or within-(w) subject factors."
}
```

Box 6.3: Example JSON representation of the `type_test` property.

``` json
{
  ...
  "type_test": "2w*2w",
  ...
}
```

Box 7.1: JSON Schema fragment defining the `test_parameters` property.

``` json
"test_parameters": {
      "type": "object",
      "description": "Test-specific parameters.",
      "properties": {
        "tail": {
          "type": "string",
          "enum": [
            "one-sided",
            "two-sided"
          ],
          "description": "Directionality of the statistical test. Required for t-tests and correlation tests."
          
        },
        "within_subject_correlation": {
          "type": "number",
          "minimum": 0,
          "maximum": 1,
          "description": "Expected correlation between repeated measurements. Required for F-tests that include within-subject factors."
        }
      }
    }
```

Box 7.3: Example JSON representation of the `test_parameters` property.

``` json
{
  ...
  "test_parameters": {
    "within_subject_correlation": 0.5
  },
  ...
}
```

### 

Box 8.1: JSON Schema representation of the `type_effect` property.

``` json
"type_effect": {
      "type": "string",
      "description": "The effect or estimand of interst.",
      "enum": [
        "main effect",
        "interaction",
        "planned contrast",
        "simple effect",
        "other"
      ]
    }
```

Box 8.2: Example JSON representation of the `type_effect` property.

``` json
{
  ...
  "type_effect": "interaction"
  ...
}
```

### Effect size

The effect size metric (e.g., correlation, Cohen's f, odds ratio) and magnitude (e.g., 0.5, 1.3) used in the power analysis should be reported (**Cris, add a part 'but are often missing REF REF REF**). Reporting only the numerical value (e.g., and effect size of 0.1) is insufficient because the same value can represent different quantities depending on the metric used (e.g., the effect size could be a Cohen's d, or a raw mean difference, or Cohen's f, or partial eta squared). Reporting the metric allows readers to determine whether the assumed effect size corresponds to the effect of interest, and increases the reproducibility of the power analysis. Reporting the magnitude allows readers to evaluate whether the assumed effect size is plausible, justified, and appropriate for the research question.

Reporting the metric may not always be sufficient trace back what it was based on. Researchers may convert the effect size they use in a power analysis from an effect size reported in another metric, calculate an effect size from summary statistics, or combine multiple effect sizes in the literature to obtain the value used in the power analysis. For example, researchers may combine information from multiple studies by averaging reported effect sizes, apply bias-adjustment methods to an effect size reported in a previous study or meta-analysis. Although this recommendation goes beyond simply reporting the power analysis, it is essential information to be able to evaluate the validity of a power analysis. Researchers should transparently report how the effect size used in the power analysis was obtained or calculated. For example, if an effect size was converted from another metric, researchers should report the original metric and the conversion formula used. If the effect size was calculated by the researchers, the calculation procedure should be described. This information allows readers to evaluate the assumptions underlying the selected effect size and improves the reproducibility of the power analysis.


Box 9.1: JSON Schema representation of the `effect_size` property.

<!--# We could also create another property to specify how the effect size was estimated where researchers can include means and standard deviations. Should we alos allow spceify the type of effect size beyond the family? For instance, Cohen's ds or drm. -->

``` json
 "effect_size": {
      "type": "object",
      "description": "The metric and magnitude of the effect size used in the power analysis.",
      "additionalProperties": false,
      "properties": {
        "metric": {
          "type": "string",
          "enum": [
            "d", 
            "g", 
            "f", 
            "r", 
            "β"
            ]
        },
        "magnitude": {
          "type": "number"
        },
        "description": {
          "type": "string",
          "description": "Additional information about how the effect size was obtained. If the effect size was converted from another metric, report the conversion formula used and its reference."
    }
      },
      "required": [
        "metric",
        "magnitude"
      ]
    }
```

Box 9.2: Example JSON representation of the `type_effect` property.

``` json
{
  ...
  "effect_size": {
    "metric": "f",
    "magnitude": 0.12,
  }
  ...
}
```

### Effect Size Justification

There are six common approaches that researchers use to justify the effect size of interest in a power analysis: effect sizes from previous studies, meta-analyses, pilot studies, Cohen's *d* thresholds, field-specific thresholds, and the smallest effect size of interest (SESOI) [@lakens2022]. Reporting this information again goes beyond the statistical properties of a power analysis, but is essential when evaluate whether the sample size justification is valid. We therefore think that in addition to reporting how the effect size was determined, researchers should provide a justification for why this effect size was used in the power analysis. 

This information is especially important to be able to detect common problems with the choice of an effect size of interest in a priori power analyses. When effect sizes are derived from previous studies or meta-analyses, researchers should explain that the effect is sufficiently similar to the planned study, as studies might differ in the population, study design, manipulation, intervention, or even outcome measures. Furthermore, researchers should reflect on the (possibly large) uncertainty in the effect size estimate of the original study. This is also true if the effect size comes from a pilot study that the researchers performed themselves. 

Instead of using an empirically observed effect size for the power analysis, researchers may rely on Cohen's *d* thresholds for small, medium, and large effects, or field-specific benchmarks. This practice can be problematic because it uses general benchmarks to design a specific study that is likely not similar to the general study in a field. Consequently, choosing an effect size solely on the basis of Cohen's *d* thresholds or other field-specific benchmarks may overlook the contextual factors that determine whether the effect size is appropriate for the planned study [@primbs_2022]. 

A better approach is to justify the effect size based on a smallest effect size of interest (SESOI), which specifies the effect a researcher does not want to miss, either because it is considered theoretically or practically meaningful for the research question. Regardless of the justification for the effect size is used to compute power, it should not simply be stated, but researchers should justify why it is a relevant effect size to design the study around. 

[**Cris, I moved part of the bias correction up - it is part of computing the effect size - but you also had: ", or use an effect size from a previous study with a simpler design to estimate an interaction effect [@langenberg2023; @sommet2023]" - but this is bad practice, right? WHy would we discussi it? **]

Box 10.1: JSON Schema representation of the `effect_size_justification` property.

``` json
"effect_size_justification": {
  "type": "object",
  "description": "Justification for the effect size estimate used in the power analysis.",
  "additionalProperties": false,
  "properties": {
    "method": {
      "type": "string",
      "enum": [
        "previous study",
        "meta-analysis",
        "pilot study",
        "Cohen threshold",
        "field-specific threshold",
        "SESOI",
        "other"
      ]
    },
    "reference": {
      "type": "string",
      "description": "Reference to the study, meta-analysis, or other source used to justify the effect size estimate."
    },
    "description": {
      "type": "string",
      "description": "Additional explanation about how the effect size was selected or calculated. For example, researchers may average effect sizes from multiple studies, use a bias-adjusted effect size or estimate an interaction effect from a independent two-group study design."
    }
  },
  "required": [
    "method"
  ]
}
```

Box 10.2: Example JSON representation of the `effect_size_justification` property.

``` json
{
 ...
  "effect_size_justification": {
    "method": "previous study",
    "reference": "https://doi.org/example/251524592097094",
    "description": "EXAMPLE?"
  }
  ...
}
```

### Sample size

The sample size used in the power analysis should be reported explicitly. Researchers should both report the total sample size, and where applicable, the sample size per group. Reporting the total and per group sample size allows the power analysis to be reproduced. The sample size can be people, or multiple observations from a single individual, or any combination thereof, and for studies with multiple observation per individual, information about the number of observations should be provided as well.

Box 11.1: JSON Schema representation of the `sample_size` property.

``` json
    "sample_size": {
      "type": "object",
      "description": "Sample size information.",
      "additionalProperties": false,
      "properties": {
        "per_group": {
          "type": "integer",
          "minimum": 1
        },
        "total": {
          "type": "integer",
          "minimum": 1
        }
      },
      "required": [
        "total"
      ]
    }
```

Box 11.2: Example JSON representation of the `sample_size` property.

``` json
{
  ...
  "sample_size": {
    "total": 186
  },
  ...
}
```

### Power

The power level should be reported because it determines the probability of detecting an effect of the specified magnitude, assuming that the effect exists. Although a power of 0.8 is commonly used as a default, researchers should explicitly report their chosen level of power, and there are good reasons to increase the statistical power 0.9 or 0.95 where possible. To facilitate machine readability and automated extraction, power should be reported as a numeric value between 0 and 1 (e.g., 0.80 rather than 80%).

[**Cris, should people report the Type 2 error or the power? **]

Box 12.1: JSON Schema representation of the `power` property.

``` json
"power": {
      "type": "number",
      "minimum": 0,
      "maximum": 1,
      "description": "Planned or achieved statistical power."
    }
```

Box 12.2: Example JSON representation of the `power` property.

``` json
{
  ...
  "power": 0.9,
  ...
}
```

### Alpha

Although 0.05 is commonly used as a default alpha level in hypothesis tests, the alpha level should be reported because it determines the threshold for statistical significance and directly affects the power of a test. Reporting the alpha level also allows peers to evaluate whether it matches the alpha level used in the study. For example, a study may adjust the alpha level to account for multiple comparisons in the results section, but use a different alpha level in the power analysis. As corrections for multiple comparisons lower the alpha level, the also lower the statistical power of the test. To facilitate machine readability and automated extraction, the alpha level should be reported as a numeric value between 0 and 1 (e.g., 0.05 rather than 5%). The justification property allows researchers to provide a justification explaining the rationale for this choice [@lakens_justify_2018], such as adjustments for multiple comparisons. [**Nice - should we say the same for power? Also justify it?**]

Box 13.1: JSON Schema representation of the `alpha` property.

``` json
 "alpha": {
  "type": "object",
  "description": "Type I error rate selected for the statistical test.",
  "additionalProperties": false,
  "properties": {
    "value": {
      "type": "number",
      "minimum": 0,
      "maximum": 1,
      "description": "Type I error rate selected for the analysis."
    },
    "justification": {
      "type": "string",
      "description": "Explanation of the rationale for selecting the alpha level. For example, researchers may adjust the alpha level to account for multiple comparisons."
    }
  },
  "required": [
    "value"
  ]
}
```

Box 13.2: Example JSON representation of the `alpha` property.

``` json
{
  ...
  "alpha": {
    "level": 0.05
    "justification": ""
  }
  ...
}
```

### Software

The software used to conduct the power analysis should be reported to allow peers to reproduce the analysis using the exact same software, and where relevant, the packages package used (e.g., the specific R package). Different software packages can return slightly different calculations for the same settings. Researchers should also provide a link to the script, notebook, or other file that contains the code used to perform the power analysis. The link is preferably a Digital Object Identifier (DOI) that links to a file in a stable data repository. If the software used is point-and-click, researchers can link to a screenshot (for example on Figshare). 

<!--# An additional property could be "version" or "code" where researchers have to include the code used (e.g., pwr.t.test(...)) -->

Box 14.1: JSON Schema representation of the `software` property.

``` json
"software": {
  "type": "object",
  "description": "Software used for the power analysis.",
  "additionalProperties": false,
  "properties": {
    "name": {
      "type": "string",
      "enum": [
        "R",
        "G*Power",
        "JASP",
        "jamovi"
      ]
    },
    "package": {
      "type": "string",
      "description": "Name of the package used to perform the power analysis."
    },,
    "file_location": {
      "type": "string",
      "format": "uri",
      "Description": "Link to the repository, file, or other online location containing the code used for the power analysis."
}
    },
  },
  "required": [
    "name"
  ]
}
```

Box 14.2: Example JSON representation of the `software` property.

``` json
{
  ...
  "software": {
    "name": "G*Power",
    "file_location":""
  }
}
```

## The JSON Structure

Box 15 shows a complete example containing all properties defined in the proposed schema.


```{r}
#| label: res_power
#| eval: FALSE ## Set to TRUE to rerun module

# Opt-out LLM
llm_use(FALSE)

# Run the module
res_power <- module_run(psych_science, "power")

# Save results
save(res_power, file = "data/res_power.RData")
```

```{r}
#| label: power_type

# Load data
load("data/res_power.RData")

# Calculate number and proportion of power analyses by type
power_type <- res_power$table |> 
  count(power_type) |> 
  mutate(
    Percent = paste0(
      round((n / sum(n)) * 100, 0),
      "%"
    )
  ) |> 
  select(
    Power = power_type, 
    Cases = n, 
    Percent
    )
```

The `power` module that we developed uses regular expressions (i.e., rule-based pattern matching) to identify sentences describing a power analysis and classifies them as a priori, sensitivity or post hoc. To identify potential power analyses, the module searches for paragraphs that satisfy three criteria: they contain the standalone word *power* (or *powers*/*powered*), at least one term associated with power analyses (e.g., *effect size*, *sample size*, *a priori*, *sensitivity*, or *post hoc*), and at least one numeric value, excluding four-digit numbers that are likely to represent publication years. Because this approach relies on regular expressions, only power analyses that explicitly include keywords such as *a priori*, *sensitivity* or *post hoc* can be classified.

To illustrate this limitation, we applied the `power` module to a sample of `r length(psych_science)` open-access articles from *Psychological Science* to estimate how many power analyses could be classified as *a priori*, *sensitivity* or *post hoc*. Across the `r length(res_power$table$power_id)` reported power analysis (although these include some double counted due to overlapping sentences), `r power_type[[4, 2]]` could not be classified, `r power_type[[1, 2]]` were classified as *a priori*, `r power_type[[3, 2]]` were classified as *sensitivity* and `r power_type[[2, 2]]` were classified as *post hoc*. These results suggest that many reported power analyses cannot be classified using the rule-based methods alone because authors often describe them without using standardized terminology. For example, researcher may write that a study was “powered at 80%”, “the desired power was .80” or “we planned to achieve an 80% probability of detecting the effect.” without explicitly stating that the analysis was *a priori*, *sensitivity* or *post hoc*.

```{r}
#| label: llm_power
#| eval: FALSE ## Set to TRUE to rerun module

# Opt-in LLM
llm_use(TRUE)

# Choose model
llm_model("ollama/qwen3.5:9b")      

# Increase the number of calls. Maximum number set to 30
llm_max_calls(2500)

# Run power module
res_power_llm <- module_run(psych_science, "power")

# Save the results
save(res_power_llm, file = "data/res_power_llm.RData")
```

```{r}
# Load data
load("data/res_power_llm.RData")

# Number of power analyses
n_power_analyses <- nrow(res_power_llm$table)
```

<!--#  Below we show the prevalence of missing information using the LLM. This was already reported in the Metacheck paper. Instead of reporting the same information twice perhaps we can show an example of poorly reported power analysis?-->

A second goal of the `power` module is to help researchers report power analyses more completely. However, extracting parameters of a power analysis from unstructured scientific text is challenging because the same information can be phrased in many different ways. Consequently, deterministic rule-based approaches fail to identify parameter information that is present but expressed in an unexpected way. To address this limitation the `power` module can optionally use an LLM to interpret flexible human language. To illustrate this approach, we applied the LLM-version of `power` module to a sample of `r length(psych_science)` open-access articles from *Psychological Science*. @tbl-power_llm shows the proportion of missing components from `r n_power_analyses` power analyses. Although the LLM-based approach can identify information expressed in a wider range of ways than rule-based methods, some parameters remain missing because they were not reported or could not be reliably extracted from the text.

```{r}
#| label: tbl-power_llm
#| tbl-cap: !expr "sprintf('Proportion of missing components from %d power analyses.', n_power_analyses)"

# Estimate the number and proportion of components from power analyses that are missing
tbl_power_llm <- res_power_llm$table |> 
  select(software_other:power_type) |> 
  mutate(
     # Combine software fields 
    software_combined = coalesce(
      software, 
      software_other
      ),
    # Combine effect size metric fields
    effect_size_metric_combined = coalesce(
      effect_size_metric,
      effect_size_metric_other
    ),
    # Combine statistical test fields
    statistical_test_combined = coalesce(
      statistical_test_other,
      statistical_test 
    )
  ) |> 
  select(
    -c(
    software, 
    software_other, 
    effect_size_metric, 
    effect_size_metric_other,
    statistical_test_other,
    statistical_test
    )
  ) |>
  mutate(
    across(
      -power_type,
      ~ is.na(.x)
    ),
    power_type = power_type == "unknown"
  ) |>
  summarise(
    across(everything(),
           list(
             n = sum,
             pcnt = mean
           )
    )
  ) |> pivot_longer(
    cols = everything()
  ) |>
  separate(
    name, 
    c("Component", "name"), 
    sep = "_(?!.*_)"
    ) |>
  pivot_wider() |>
  mutate(pcnt = sprintf("%d%%", round(pcnt*100)))

# Display the table
tbl_power_llm |> 
  flextable() |>
  align(
    align = "right", 
    j = "pcnt", 
    part = "all"
    ) |>
  set_header_labels(
    n = "Missing",
    pcnt = "Percent"
    ) |>
  width(
    1:3, 
    c(2, 1, 1)
    )
```

Although LLMs provide a flexible way to parse natural language variation, they treat the symptoms of inconsistent reporting rather than its root cause. When information is inherently ambiguous or incomplete, an LLM may misclassify a power analysis or extract incorrect parameters. Moreover, using LLMs for routine information-extraction tasks incurs substantial computational and energy costs. A preferable long-term solution is to reduce the need for LLM-based approaches by making scientific text easier to extract in the first place. This goal is best achieved through standardized reporting conventions.

#### Validation

The performance of the `power` module was evaluated on `r n_papers` open-access articles published in Psychological Science, for which a ground-truth dataset was manually created. Accuracy was highest for identifying the alpha level (`r alpha_acc`), software used for the analysis (`r software_acc`) and achieved power (`r power_acc`). Accuracy was also relatively high for extracting the expected effect size (`r es_acc`) and sample size (`r sample_acc`), and moderate for identifying the statistical test (`r test_acc`). The lowest accuracy was observed for the type of power analysis (`r type_acc`) and the metric used to express the expected effect size (`r esmetric_acc`). These results indicate that the module performs better for components that are typically reported using standardized numerical values or commonly recognized software names, whereas components that depend more heavily on contextual interpretation or inconsistent reporting terminology are more challenging to extract accurately.

#### Machine-readable power analyses

Akin to the concept of machine-readable hypothesis [@lakens_2021], one possible solution would be to represent power analyses directly in a machine-readable format, such as JavaScript Object Notation (JSON). For example, the `power` module uses a JSON schema that defines the structure of a power analysis report and guides the LLM in extracting and organizing relevant information. A schema can specify which fields are required, which values are permitted, and how the information should be structured, making the resulting data easier for software to process. Such a representation could include predefined fields for the type of power analysis, statistical test, sample size, alpha level, target power, effect size, and other relevant parameters, such as the justification for the assumed effect size \[\@lakens2022\].

However, requiring researchers to produce and submit power analyses directly as JSON files is currently unlikely to be practical. Although such a format would facilitate automated processing and validation, it would shift the burden of structuring information from automated systems to researchers and create an additional reporting requirement. A more feasible approach may be to maintain human-readable reporting formats while adopting standardized structures that allow power analysis information to be converted into machine-readable representations.


## Standardized reporting

Now that we have defined a JSON Schema that specifies which information should be reported and how it should be structured, the next step is to establish standardized reporting guidelines for power analysis. These guidelines should ensure that information is interpretable by human readers but also structured in a way that automated tools like Metacheck can extract using deterministic approaches (e.g., regular expressions). The extracted information can then be used to populate a JSON object, which can be validated against the JSON Schema to ensure that all required fields are present and that reported values conform to the specified constraints.

<!--# The way I conceive this is that the Metacheck power module will use rule-based code to extract this information, represent it as a structured object, and validate it against the JSON schema. -->



## Discussion

The proposed schema represents an initial step toward a standardized representation of power analyses that can improve transparency by imposing a consistent structure on the information required to evaluate and reproduce power analyses, while also facilitating information extraction using rule-based automated tools.

The current version focuses on commonly used statistical tests, including correlations, *t*-tests, and *F*-tests. Extending the standard to accommodate more complex analytical approaches, such as power analyses for mixed-effects models or simulation-based power analyses, will require identifying additional parameters and reporting requirements. Developing a comprehensive standard that covers the diversity of power analysis approaches used across disciplines will therefore require a community effort involving researchers and other stakeholders, such as journals. Such collaborative development can ensure that future extensions of the schema reflect the needs of different research communities while maintaining a consistent structure that supports both human interpretation and automated extraction.

Although structured formats such as JSON provide clear advantages for automated processing and validation, requiring researchers to manually complete structured objects for each power analysis would introduce additional reporting demands and may reduce adoption. The proposed approach therefore aims to achieve a compromise between machine readability and practical feasibility by maintaining conventional human-readable reporting while introducing sufficient structure for automated extraction and validation. Nevertheless, future developments could explore workflows in which software generates machine-readable representations directly, reducing the need for manual data entry while preserving the advantages of structured reporting.

Standardizing the way information is reported in scientific articles has benefits beyond improving the transparency of individual power analyses. When information is consistently structured across articles, automated tools can more reliably extract and organize reported information into databases containing standardized records of scientific practices. Such databases would support the FAIR principles [@wilkinson_2016] by improving the findability of relevant information and facilitating the reuse of scientific information for large-scale assessments. For example, researchers could identify studies that meet specific methodological criteria, include participants from particular populations, or evaluate the same intervention. By making scientific information easier to extract, organize, and query, standardized reporting conventions can improve not only transparency but also the discoverability and reuse of research outputs.

```{r}
#| label: session-info
#| include: false

capture.output(
  sessioninfo::session_info(),
  file = "session-info.txt"
)
```

## References
